Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-11-22 Paper/ms.tex
Data in latex comment
Minor fixes

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Recently, it has become clear that RV data
can contain periodic signals due to stellar activity, which can be difficult
to distinguish from planetary signals. However, even doing the inference
under a given set of siplifying assumptions (e.g. no stellar activity) can
be difficult. It is common for the posterior distribution for the
planet parameters, such as orbital periods,
to be multimodal and to have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence), as a
function of $N$, is required. Rather than doing separate runs with different
values of $N$, we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run, which takes $\sim$ 15 minutes on a typical modern desktop computer.
We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
for additional signals in
$\nu$ Oph with periods of 36.12 $\pm$ 0.04 days and
75.57 $\pm$ 0.76 days; the posterior probability that at least one of these
exists is 79\%. The results also suggest Gliese 581 hosts many (7-15) ``planets''
(or other causes of other periodic
signals), but only four have well determined periods. The analysis of both
of these datasets shows phase transitions exist which are difficult to
negotiate without Nested Sampling.
\end{abstract}

@@

\section{Introduction}
The number of known extrasolar planets has exploded in the last two
decades, driven by improvements in
all of the different techniques used to detect and characterise exoplanets,
including the radial velocity (RV) method \citep[e.g.][]{2012PASJ...64..135S},
the transit method \citep[e.g.]{2014PNAS..11112647B},
and gravitational microlensing
\citep[e.g.][]{2014ApJ...785..155B, 2014ApJ...790...14Y}.

@@

>>>>

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Recently, it has become clear that RV data
can contain periodic signals due to stellar activity, which can be difficult
to distinguish from planetary signals. However, even doing the inference
under a given set of simplifying assumptions (e.g. no stellar activity) can
be difficult. It is common for the posterior distribution for the
planet parameters, such as orbital periods,
to be multimodal and to have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence), as a
function of $N$, is required. Rather than doing separate runs with different
values of $N$, we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run, which takes $\sim$ 15 minutes on a typical modern desktop computer.
We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
for additional signals in
$\nu$ Oph with periods of 36.12 $\pm$ 0.04 days and
75.57 $\pm$ 0.76 days; the posterior probability that at least one of these
exists is 79\%. The results also suggest Gliese 581 hosts many (7-15) ``planets''
(or other causes of other periodic
signals), but only four have well determined periods. The analysis of both
of these datasets shows phase transitions exist which are difficult to
negotiate without Nested Sampling.
\end{abstract}

@@

\section{Introduction}
The number of known extrasolar planets has exploded in the last two
decades. This has been driven by improvements in
all of the different techniques used to detect and characterise exoplanets,
including the radial velocity (RV) method \citep[e.g.][]{2012PASJ...64..135S},
the transit method \citep[e.g.]{2014PNAS..11112647B},
and gravitational microlensing
\citep[e.g.][]{2014ApJ...785..155B, 2014ApJ...790...14Y}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DATA USED

% Nu Oph
% Time (days), Radial velocity (m/s), Radial velocity errorbar (m/s)
%2.337364819999999781e+03 -2.061999999999999886e+02 4.299999999999999822e+00
%2.487101369999999861e+03 2.615000000000000000e+02 4.200000000000000178e+00
%2.507077150000000074e+03 3.403000000000000114e+02 3.799999999999999822e+00
%2.541955269999999928e+03 4.198999999999999773e+02 5.799999999999999822e+00
%2.541958299999999781e+03 4.293999999999999773e+02 4.000000000000000000e+00
%2.736306950000000143e+03 -2.519999999999999929e+01 5.299999999999999822e+00
%2.857037139999999908e+03 -2.150000000000000000e+01 5.000000000000000000e+00
%3.100276800000000094e+03 5.113000000000000114e+02 3.700000000000000178e+00
%3.499178260000000137e+03 1.225000000000000000e+02 5.299999999999999822e+00
%3.636025169999999889e+03 3.305000000000000000e+02 6.200000000000000178e+00
%3.963951230000000123e+03 -2.350999999999999943e+02 4.900000000000000355e+00
%4.309113010000000031e+03 -2.686999999999999886e+02 5.200000000000000178e+00
%4.338994730000000345e+03 -3.363999999999999773e+02 3.399999999999999911e+00
%4.560239679999999680e+03 -1.160000000000000000e+02 3.500000000000000000e+00
%4.588269440000000031e+03 -2.230000000000000071e+01 5.000000000000000000e+00
%4.674023040000000037e+03 1.691999999999999886e+02 4.099999999999999645e+00
%4.674025380000000041e+03 1.686999999999999886e+02 4.099999999999999645e+00
%4.677067339999999604e+03 1.716999999999999886e+02 3.799999999999999822e+00
%4.928330149999999776e+03 -3.908000000000000114e+02 3.500000000000000000e+00
%4.952301330000000235e+03 -3.595000000000000000e+02 4.500000000000000000e+00
%4.953310220000000299e+03 -3.531999999999999886e+02 4.099999999999999645e+00
%5.132930250000000342e+03 7.940000000000000568e+01 6.700000000000000178e+00
%5.267333859999999731e+03 1.450000000000000000e+02 3.500000000000000000e+00
%5.296329950000000281e+03 4.579999999999999716e+01 3.500000000000000000e+00
%5.324284969999999703e+03 -3.520000000000000284e+01 4.000000000000000000e+00
%5.347153669999999693e+03 -1.170000000000000000e+02 4.099999999999999645e+00
%5.384187179999999898e+03 -1.946999999999999886e+02 4.200000000000000178e+00
%5.398047959999999875e+03 -2.165999999999999943e+02 3.700000000000000178e+00
%5.436026149999999689e+03 -2.645000000000000000e+02 3.500000000000000000e+00
%5.467928210000000036e+03 -2.518000000000000114e+02 3.700000000000000178e+00
%5.613381190000000061e+03 6.979999999999999716e+01 3.500000000000000000e+00
%5.656341529999999693e+03 2.221999999999999886e+02 3.299999999999999822e+00
%5.657340799999999945e+03 2.265000000000000000e+02 3.799999999999999822e+00
%5.690210130000000390e+03 3.385000000000000000e+02 6.000000000000000000e+00
%5.712159810000000107e+03 4.010000000000000000e+02 3.799999999999999822e+00
%5.763070190000000366e+03 3.993000000000000114e+02 3.399999999999999911e+00
%5.766075829999999769e+03 3.908999999999999773e+02 3.600000000000000089e+00
%5.770066890000000058e+03 3.886000000000000227e+02 4.000000000000000000e+00
%5.784950219999999717e+03 3.693000000000000114e+02 4.099999999999999645e+00
%5.787005500000000211e+03 3.641999999999999886e+02 3.600000000000000089e+00
%5.788002169999999751e+03 3.663000000000000114e+02 3.700000000000000178e+00
%5.804928630000000339e+03 2.938000000000000114e+02 3.799999999999999822e+00
%5.809972679999999855e+03 2.825000000000000000e+02 3.700000000000000178e+00
%5.846919119999999566e+03 1.640999999999999943e+02 4.900000000000000355e+00

% Gliese 581
% First 122 points are HIRES, then HARPS follows
%1409.76222   6.96 1.89 
%1586.14605 -10.24 3.22 
%1704.91213   0.47 2.89 
%2003.95507  -4.37 3.65 
%2100.86678 -19.45 2.22 
%2161.73096   3.35 2.19 
%2162.73165  10.73 2.38 
%2335.15024   6.87 2.33 
%2487.79326  -7.30 3.77 
%2712.05433 -10.96 2.10 
%2828.87708   7.75 2.25 
%3077.15398  -7.18 3.18 
%3239.82508  -2.50 2.60 
%3426.14551  -2.74 4.82 
%3748.16323  -4.97 1.96 
%3754.14596  -8.68 1.97 
%3932.80588   2.40 1.95 
%3960.83771  12.14 1.90 
%3961.77279   4.38 1.57 
%3962.77418  -4.97 2.40 
%3963.81985  -9.92 1.78 
%3982.78069  14.04 1.92 
%4130.14636 -14.79 1.84 
%4131.16220  -4.58 2.58 
%4139.16155   1.60 1.92 
%4246.81907  -4.50 1.18 
%4247.92767  -9.21 1.84 
%4248.82815  -7.84 1.48 
%4249.90281   6.45 1.38 
%4250.81292  18.17 1.95 
%4251.81942   7.61 1.30 
%4255.79771   7.35 3.62 
%4277.76996  17.11 1.68 
%4278.78138  14.06 1.71 
%4279.78247 -10.60 2.03 
%4285.78969 -15.08 2.54 
%4294.84381  -0.04 2.04 
%4300.81847  -0.76 1.83 
%4304.79872  16.63 2.24 
%4305.80545   2.61 1.45 
%4306.83201  -8.05 2.31 
%4307.83943  -8.03 1.47 
%4308.82016   1.74 1.27 
%4309.80568  13.32 1.26 
%4310.80271  11.19 1.26 
%4311.79457  -4.40 1.23 
%4312.79306  -7.14 1.13 
%4313.80176   8.14 1.45 
%4314.82230  13.67 1.35 
%4318.79357  -5.91 1.44 
%4335.75166   5.83 1.35 
%4336.78425  14.68 1.84 
%4339.72676  -8.58 2.69 
%4343.74861  -0.29 2.06 
%4345.74397  -5.82 1.94 
%4461.17458   3.57 2.13 
%4545.09518   4.98 1.95 
%4546.07813  17.48 1.52 
%4547.07715   9.06 2.74 
%4547.99454  -3.97 2.45 
%4600.98896  13.54 2.03 
%4601.90368  -8.33 2.69 
%4633.86049  -6.47 1.49 
%4634.86134 -12.30 1.27 
%4635.87975  -5.69 1.19 
%4636.88690  12.26 2.07 
%4637.88502  15.38 1.56 
%4638.94288  -1.08 1.40 
%4639.91382 -11.27 1.39 
%4640.92943  -6.50 1.95 
%4641.90267   5.64 1.33 
%4643.88456   0.95 1.74 
%4666.84246  -7.60 1.25 
%4667.84194  -8.11 2.07 
%4671.85435  -9.71 1.33 
%4672.83453  -3.24 1.00 
%4673.84973   8.11 1.03 
%4686.82794   2.60 2.22 
%4688.77854  -9.57 1.96 
%4689.77880   3.13 1.29 
%4701.79211  17.51 1.17 
%4702.81250  11.35 1.72 
%4703.80853  -1.73 1.54 
%4704.80341  -7.99 1.20 
%4717.72887  14.81 1.48 
%4718.73784   5.10 2.03 
%4719.73866  -4.93 1.86 
%4720.73430 -14.79 1.46 
%4721.73009  -9.86 1.33 
%4722.74036   3.80 1.58 
%4723.73908   6.85 1.62 
%4903.15067 -14.41 2.08 
%5021.91811 -10.11 1.43 
%5022.91286   1.39 1.86 
%5024.95012   9.32 1.57 
%5049.86010   4.91 0.90 
%5050.80706  13.45 0.79 
%5051.84043   5.28 1.17 
%5052.86710  -8.12 2.09 
%5053.79905 -10.89 0.80 
%5201.15720  16.10 1.58 
%5202.15734   9.80 1.61 
%5203.14807  -8.37 1.53 
%5258.15584   2.44 2.22 
%5260.13379  16.54 2.11 
%5338.81312  -4.84 0.53 
%5338.90357  -3.10 1.59 
%5338.98795  -1.62 1.33 
%5339.06181  -1.56 1.82 
%5340.05371   9.91 1.63 
%5340.86600  11.62 1.38 
%5340.95260   9.87 0.58 
%5341.03351   6.69 0.76 
%5341.79524  -1.63 0.80 
%5342.00395  -1.67 0.88 
%5342.06383  -5.26 1.46 
%5369.93610 -15.90 1.40 
%5370.87580 -14.21 1.48 
%5371.93429   6.73 1.13 
%5408.89194  -0.40 2.00 
%5409.76684  15.21 1.48 
%5409.86682  15.68 1.98 
%3.152712894000113010e+3 -1.076949579832486847e+01 1.100000000000000089e+00
%3.158663459999952465e+03 -1.927949579832421989e+01 1.199999999999999956e+00
%3.511773341000080109e+03 -7.469495798325453961e+00 1.199999999999999956e+00
%3.520744746000040323e+03 9.360504201675112768e+00 1.300000000000000044e+00
%3.573512039000168443e+03 -7.194957983251981659e-01 1.199999999999999956e+00
%3.574522328999824822e+03 8.800504201674996807e+00 1.100000000000000089e+00
%3.575480748999863863e+03 3.360504201674885394e+00 1.000000000000000000e+00
%3.576536046000197530e+03 -7.339495798325046394e+00 1.000000000000000000e+00
%3.577592602999880910e+03 -1.124949579832446034e+01 1.199999999999999956e+00
%3.578510712999850512e+03 -8.694957983248485789e-01 9.000000000000000222e-01
%3.578629602000117302e+03 1.740504201675818763e+00 1.100000000000000089e+00
%3.579462557000108063e+03 1.217050420167531399e+01 9.000000000000000222e-01
%3.579621048999950290e+03 1.388050420167452614e+01 1.000000000000000000e+00
%3.585461769999936223e+03 5.590504201675727813e+00 1.000000000000000000e+00
%3.586465158000122756e+03 -3.939495798324088582e+00 6.999999999999999556e-01
%3.587464703999925405e+03 -1.753949579832436712e+01 1.399999999999999911e+00
%3.588538062000181526e+03 -8.759495798324579141e+00 2.299999999999999822e+00
%3.589462024000007659e+03 4.800504201675437344e+00 6.999999999999999556e-01
%3.590463895000051707e+03 1.240050420167548850e+01 8.000000000000000444e-01
%3.591466483999975026e+03 6.640504201675057061e+00 6.999999999999999556e-01
%3.592464813000056893e+03 -6.219495798324814473e+00 8.000000000000000444e-01
%3.606551678999792784e+03 1.405050420167519576e+01 1.899999999999999911e+00
%3.607507526999805123e+03 9.530504201675782383e+00 1.000000000000000000e+00
%3.608482642999850214e+03 -4.709495798325136207e+00 1.199999999999999956e+00
%3.609488452000077814e+03 -1.422949579832533118e+01 1.500000000000000000e+00
%3.757877319000195712e+03 3.670504201675584000e+00 1.000000000000000000e+00
%3.760875475000124425e+03 -3.069495798325050373e+00 1.300000000000000044e+00
%3.761859215999953449e+03 6.250504201675610716e+00 1.300000000000000044e+00
%3.811846942999865860e+03 4.690504201676048979e+00 1.199999999999999956e+00
%3.813827016999945045e+03 -1.071949579832498500e+01 9.000000000000000222e-01
%3.830836957000195980e+03 -2.109495798324090288e+00 9.000000000000000222e-01
%3.862701441000215709e+03 -5.994957983244120214e-01 9.000000000000000222e-01
%3.864713661999907345e+03 1.355050420167458469e+01 1.000000000000000000e+00
%3.867752171000000089e+03 -9.449495798325102669e+00 1.100000000000000089e+00
%3.870696603000164032e+03 5.030504201675611853e+00 1.100000000000000089e+00
%3.882657763000112027e+03 -1.153949579832413974e+01 9.000000000000000222e-01
%3.887690737999975681e+03 -5.299495798324116436e+00 9.000000000000000222e-01
%3.918621751000173390e+03 7.230504201675813647e+00 1.100000000000000089e+00
%3.920594947000034153e+03 -1.805949579832422103e+01 1.000000000000000000e+00
%3.945543122000060976e+03 9.820504201675461786e+00 1.000000000000000000e+00
%3.951485927000176162e+03 5.480504201674563092e+00 9.000000000000000222e-01
%3.975471595999784768e+03 -7.649495798323968643e+00 1.000000000000000000e+00
%3.979543974999804050e+03 -8.039495798325191345e+00 1.199999999999999956e+00
%4.166874181999824941e+03 -1.481949579832431141e+01 1.100000000000000089e+00
%4.170853961000218987e+03 7.430504201675347531e+00 9.000000000000000222e-01
%4.194872349000070244e+03 -1.367949579832483664e+01 1.100000000000000089e+00
%4.196750384000129998e+03 1.457050420167504967e+01 1.199999999999999956e+00
%4.197845038999803364e+03 1.538050420167458299e+01 1.199999999999999956e+00
%4.198855508999899030e+03 -3.269495798324584257e+00 1.199999999999999956e+00
%4.199732869000174105e+03 -5.909495798324115867e+00 1.000000000000000000e+00
%4.200910914999898523e+03 -2.694957983244705702e-01 1.000000000000000000e+00
%4.201868553999811411e+03 9.050504201676190519e+00 1.000000000000000000e+00
%4.202882594999857247e+03 1.327050420167452671e+01 1.000000000000000000e+00
%4.228741560999769717e+03 8.380504201674909837e+00 1.100000000000000089e+00
%4.229700478000100702e+03 1.080050420167566472e+01 1.399999999999999911e+00
%4.230762136000208557e+03 -1.799495798325168039e+00 1.000000000000000000e+00
%4.234645916000008583e+03 1.381050420167539983e+01 1.100000000000000089e+00
%4.253633167999796569e+03 -9.309495798325073679e+00 1.000000000000000000e+00
%4.254664809999987483e+03 -4.359495798324175553e+00 1.000000000000000000e+00
%4.291568849999923259e+03 -7.179495798323998201e+00 1.300000000000000044e+00
%4.292590813999995589e+03 -4.949579832391748369e-02 9.000000000000000222e-01
%4.293625868000090122e+03 9.980504201674733622e+00 1.000000000000000000e+00
%4.295639448000118136e+03 -1.049949579832443192e+01 1.100000000000000089e+00
%4.296606114000082016e+03 -2.112949579832523739e+01 1.199999999999999956e+00
%4.297641939000226557e+03 -7.539495798324580278e+00 1.000000000000000000e+00
%4.298567600000184029e+03 7.790504201675929608e+00 1.100000000000000089e+00
%4.299622194999828935e+03 1.060050420167613083e+01 1.500000000000000000e+00
%4.300619109999854118e+03 1.005042016757329293e-01 1.000000000000000000e+00
%4.315507493999786675e+03 1.331050420167478876e+01 1.600000000000000089e+00
%4.317480847000144422e+03 -8.449495798323880535e+00 1.000000000000000000e+00
%4.319490526999812573e+03 2.190504201674770002e+00 1.300000000000000044e+00
%4.320544071999844164e+03 1.182050420167612970e+01 1.000000000000000000e+00
%4.323507050000131130e+03 -1.125949579832408176e+01 3.200000000000000178e+00
%4.340555780999828130e+03 -1.549495798323974327e+00 9.000000000000000222e-01
%4.342486204000189900e+03 1.844050420167597792e+01 1.000000000000000000e+00
%4.349515163000207394e+03 -1.179949579832495488e+01 1.000000000000000000e+00
%4.530855659999884665e+03 6.150504201675843774e+00 1.000000000000000000e+00
%4.550831273999996483e+03 6.140504201676222351e+00 9.000000000000000222e-01
%4.553803722000215203e+03 -1.330949579832463314e+01 8.000000000000000444e-01
%4.563838002000004053e+03 -3.339495798325486930e+00 9.000000000000000222e-01
%4.566761146999895573e+03 1.505042016756164003e-01 1.100000000000000089e+00
%4.567791670999955386e+03 8.300504201676162097e+00 1.000000000000000000e+00
%4.569793302000034600e+03 -1.422949579832533118e+01 1.000000000000000000e+00
%4.570804248000029474e+03 -1.572949579832538802e+01 1.000000000000000000e+00
%4.571818376000039279e+03 3.530504201675555009e+00 1.100000000000000089e+00
%4.587861967999953777e+03 2.970504201675439049e+00 1.500000000000000000e+00
%4.588838798999786377e+03 8.170504201675754530e+00 1.100000000000000089e+00
%4.589827492999844253e+03 5.160504201676019420e+00 1.100000000000000089e+00
%4.590819633999839425e+03 -5.959495798323999338e+00 1.000000000000000000e+00
%4.591817119999788702e+03 -2.045949579832395671e+01 1.500000000000000000e+00
%4.592827337000053376e+03 -8.829495798325481815e+00 9.000000000000000222e-01
%4.610742932000197470e+03 1.883050420167542427e+01 1.100000000000000089e+00
%4.611713477000128478e+03 8.080504201675609011e+00 9.000000000000000222e-01
%4.616713028000202030e+03 8.550504201675579452e+00 1.300000000000000044e+00
%4.639686739000026137e+03 -1.078949579832411132e+01 1.100000000000000089e+00
%4.640654409000184387e+03 -1.055949579832393681e+01 1.300000000000000044e+00
%4.641631705999840051e+03 2.030504201675498166e+00 1.000000000000000000e+00
%4.643644998000003397e+03 2.210504201675789204e+00 1.300000000000000044e+00
%4.644587028000038117e+03 -1.027949579832387883e+01 1.199999999999999956e+00
%4.646625357000157237e+03 -7.129495798324114730e+00 1.199999999999999956e+00
%4.647579119000118226e+03 7.150504201675289551e+00 1.100000000000000089e+00
%4.648484817000105977e+03 9.600504201674908700e+00 1.100000000000000089e+00
%4.661553706000093371e+03 -1.198949579832486734e+01 1.199999999999999956e+00
%4.662549407999962568e+03 -7.094957983238003862e-01 1.300000000000000044e+00
%4.663544865999836475e+03 1.364050420167473021e+01 1.100000000000000089e+00
%4.664553042000159621e+03 1.352050420167572042e+01 1.199999999999999956e+00
%4.665569376000203192e+03 5.950504201674533533e+00 1.000000000000000000e+00
%4.672531723000109196e+03 -5.569495798324552993e+00 1.800000000000000044e+00
%4.674524120000191033e+03 9.440504201675636864e+00 1.300000000000000044e+00
%4.677505113999824971e+03 -8.939495798324870179e+00 1.100000000000000089e+00
%4.678556785000022501e+03 1.560504201675527725e+00 1.100000000000000089e+00
%4.679504033999983221e+03 1.368050420167499226e+01 1.500000000000000000e+00
%4.681514142999891192e+03 3.410504201674768865e+00 1.399999999999999911e+00
%4.682503343000076711e+03 -7.349495798324667817e+00 1.300000000000000044e+00
%4.701485073999967426e+03 1.412050420167609843e+01 1.199999999999999956e+00
%4.703513042000122368e+03 -2.069495798323828240e+00 1.199999999999999956e+00
%4.708479050000198185e+03 -4.589495798324350062e+00 1.199999999999999956e+00
%4.721473032999783754e+03 -1.558949579832535903e+01 1.199999999999999956e+00
%4.722472370999865234e+03 2.590504201675614127e+00 1.100000000000000089e+00
